Insect venom anaphylaxis: a case - control study of the European Anaphylaxis Registry

### 2019-07-24 analysis/paper/paper.Rmd
printout prof worm

The data show that `r {{rdb$q_340_insects=="yellow jacket"} %>% sum(na.rm=T)/sum({!is.na(rdb$q_340_insects)})*100} %>% {round(.,1)}`% of all insect elicited cases were elicited by yellow jackets, followed by bees (`r {{rdb$q_340_insects=="bee"} %>% sum(na.rm=T)/sum({!is.na(rdb$q_340_insects)})*100} %>% {round(.,1)}`%). The insect venom elicited cases occurred mostly in outdoor places (`r rdb %>% filter(!is.na(q_340_insects)) %>% group_by(q_152_location) %>% summarize(n=n()) %>% {.[c(4),2] %>% sum()/(.[,2] %>% sum)*100} %>% round(1)`%) patients' homes (`r rdb %>% filter(!is.na(q_340_insects)) %>% group_by(q_152_location) %>% summarize(n=n()) %>% {.[c(8),2] %>% sum()/(.[,2] %>% sum)*100} %>% round(1)`%) or urban places (`r rdb %>% filter(!is.na(q_340_insects)) %>% group_by(q_152_location) %>% summarize(n=n()) %>% {.[c(5),2] %>% sum()/(.[,2] %>% sum)*100} %>% round(1)`%).

@@

Recent expert position on the future anaphylaxis research indicates the need for a more precise description of diagnosis, biomarkers, and phenotypes of anaphylaxis [@JimenezRodriguez2018]. Nevertheless, in order to facilitate precision-medicine approach [@Muraro2017] in the treatment of anaphylaxis - we first need to investigate its subtypes and pathomechanisms better.

@@

This study aimed to identify current patterns of Insect venom anaphylaxis (IVA) in symptoms, cofactors, and management and compare them to other types of anaphylaxis (non-IVA) in a case-control manner.

@@

The cases were either assigned to IVA group (if the triggering factor was an insect sting) or non-IVA group (other elicitors of anaphylaxis).
We compared the frequency of various symptoms, cofactors, known to increase the risk of severe anaphylaxis, [@Worm2018] and management in both groups.

Due to a large number of documented reactions in the European Anaphylaxis registry - we were able to match the IVA cases to selected non-IVA cases according to sex and age to reduce the comparison bias. Subsequently, we compared the management in both groups and matched the control group according to the severity of a reaction. Propensity score matching has been performed using the "MatchIt" package for R [@matchit]. The results of the propensity score matching are in Fig. \ref(fig:flow)B-D.

We trained a random forest classifier (using the "randomForest" package for R [@rf]) in order to find which therapy approaches were best fit for classifying cases either to IVA or non-IVA group and presented the results as mean decrease in Gini. We also performed an association study of therapeutic interventions and symptoms. The resulting phi values are presented in a heatmap with automatic clustering using Ward's D and euclidean distancing method [@heatmaply].

@@

## Symptoms
IVA showed a specific symptom pattern. Patients, who underwent IVA, more often experienced cardiologic symptoms (dizziness, reduced alertness, unconsciousness) than in other elicitors of anaphylaxis and less often showed gastrointestinal symptoms (Fig \@ref(fig:figsymptoms)A). The difference in severe hypotension frequency was especially prominent in the younger group under 22 (Fig. S\@ref(fig:supplement)).

@@

Severe reactions were more prevalent in older patients in comparison to patients under 22, and in IVA cases vs. other elicitors (Fig. \@ref(fig:sev)D). There were no differences in severity of reactions elicited by yellow-jackets and insect species.

@@

In general `r treact[1,2] %>% round(1)`% of patients with insect allergy had experienced anaphylaxis in the past. If the reaction was elicited by other elcitors previous reactions were more frequently seen (`r treact[2,2]%>% round(1)`%, `r p %>% pval()`).  We observed `r sum(repeated_reaction_in_Reg)` patients with two documented reactions in our registry. Out of these `r sum(greater_same[,2])` (`r (sum(greater_same[,2])/sum(repeated_reaction_in_Reg)*100)%>% round(1)`%) had Insect elicited anaphylaxis and in `r greater_same[1,2]` (`r greater_same[1,2]/sum(greater_same[,2])*100`%) the following reaction was more severe than before. In `r greater_same[3,2]` (`r (greater_same[3,2]/sum(greater_same[,2])*100) %>% round(1)`%) cases the reaction was graded exactly as before.

>>>>

The data showed that `r {{rdb$q_340_insects=="yellow jacket"} %>% sum(na.rm=T)/sum({!is.na(rdb$q_340_insects)})*100} %>% {round(.,1)}`% of all insect elicited cases were elicited by yellow jackets, followed by bees (`r {{rdb$q_340_insects=="bee"} %>% sum(na.rm=T)/sum({!is.na(rdb$q_340_insects)})*100} %>% {round(.,1)}`%). The insect venom elicited cases occurred mostly in outdoor places (`r rdb %>% filter(!is.na(q_340_insects)) %>% group_by(q_152_location) %>% summarize(n=n()) %>% {.[c(4),2] %>% sum()/(.[,2] %>% sum)*100} %>% round(1)`%) patients' homes (`r rdb %>% filter(!is.na(q_340_insects)) %>% group_by(q_152_location) %>% summarize(n=n()) %>% {.[c(8),2] %>% sum()/(.[,2] %>% sum)*100} %>% round(1)`%) or urban places (`r rdb %>% filter(!is.na(q_340_insects)) %>% group_by(q_152_location) %>% summarize(n=n()) %>% {.[c(5),2] %>% sum()/(.[,2] %>% sum)*100} %>% round(1)`%).

@@

Recent expert position on the future of anaphylaxis research indicates the need for a more precise description of diagnosis, biomarkers, and phenotypes of anaphylaxis [@JimenezRodriguez2018]. Nevertheless, in order to facilitate precision-medicine approach [@Muraro2017] in the treatment of anaphylaxis - we first need to investigate its subtypes and pathomechanisms better.

@@

This study aimed to identify current patterns of insect venom anaphylaxis (IVA) in regard to symptoms, cofactors, and management and compare them to other types of anaphylaxis (non-IVA) in a case-control manner.

@@

The cases were either assigned to IVA group (if the triggering factor was an insect sting) or non-IVA group (other elicitors of anaphylaxis).
We compared the frequency of various symptoms, cofactors --- known to increase the risk of severe anaphylaxis, [@Worm2018] and management in both groups.

Due to a large number of documented reactions in the European Anaphylaxis registry - we were able to match the IVA cases to selected non-IVA cases according to sex and age to reduce the comparison bias. Subsequently, we compared the management in both groups and matched the control group according to the severity of a reaction. Propensity score matching has been performed using the "MatchIt" package for R [@matchit]. The results of the propensity score matching are in Fig. \@ref(fig:flow)B-D.

We trained a random forest classifier (using the "randomForest" package for R [@rf]) in order to find therapy approaches that varied the most between IVA / non-IVA group and presented the results Gini importance [@Strobl2007]. We also performed an association study of therapeutic interventions and symptoms. The resulting phi values are presented in a heatmap with automatic clustering using Ward’s Agglomerative Hierarchical Clustering Method with Euclidean distance [@heatmaply].

@@

## Symptoms
IVA showed a specific symptom pattern. Patients, who underwent IVA, more often experienced cardiologic symptoms (dizziness, reduced alertness, unconsciousness) than in other elicitors of anaphylaxis and less often showed skin or gastrointestinal symptoms (Fig \@ref(fig:figsymptoms)A). The difference in severe hypotension frequency was especially prominent in the younger group under 22 (Fig. S\@ref(fig:supplement)).

@@

Severe reactions were more prevalent in older patients in comparison to patients under 22, and in IVA cases vs. other elicitors (Fig. S\@ref(fig:sev)D). There were no differences in severity of reactions elicited by yellow-jackets and other insect species.

@@

In general `r treact[1,2] %>% round(1)`% of patients with insect allergy had experienced anaphylaxis in the past. If the reaction was elicited by other elcitors previous reactions were more frequently seen (`r treact[2,2]%>% round(1)`%, `r p %>% pval()`).  We observed `r sum(repeated_reaction_in_Reg)` patients with two documented reactions in our registry. Out of these `r sum(greater_same[,2])` (`r (sum(greater_same[,2])/sum(repeated_reaction_in_Reg)*100)%>% round(1)`%) had Insect elicited anaphylaxis and in `r greater_same[1,2]` (`r (greater_same[1,2]/sum(greater_same[,2])*100 )%>%round(1) `%) the following reaction was more severe than before. In `r greater_same[3,2]` (`r (greater_same[3,2]/sum(greater_same[,2])*100) %>% round(1)`%) cases the reaction was graded exactly as before.
